---
title: "WL2_2025_Germination"
output: html_document
date: "2025-05-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(googlesheets4)
library(dplyr)
library(tidyverse)
library(emmeans)
library(lmtest)
```

### Load data

```{r}
url <- "https://docs.google.com/spreadsheets/d/1WP4QA1Z4SRlhaBa_PJOcZQvDXj2JjPd-z82jmw2eirY/edit?gid=0#gid=0"
germ <- read_sheet(url, sheet="Both_Weeks")
```

#### Update seeds_per_pot for rows where Num_germ is greater

```{r}
germ_update_planted_count <- germ %>%
  mutate(seeds_per_pot = if_else(Unique.ID %in% c(1842, 1873, 1967, 2345, 2079, 2250) & seeds_per_pot == 3, 4, seeds_per_pot)) %>%
  mutate(seeds_per_pot = if_else(Unique.ID %in% c(1839) & seeds_per_pot == 3, 5, seeds_per_pot)) %>%
  mutate(seeds_per_pot = if_else(Unique.ID %in% c(2091) & seeds_per_pot == 1, 2, seeds_per_pot))
```

#### Pivot long and convert date to days to germ

```{r}
#remove Num_germ and Notes columns for now
germ_update_planted_count_no_cum_germ <- germ_update_planted_count[,c(1:7,9:24)]

#pivot long
germ_long <- germ_update_planted_count_no_cum_germ %>%
  pivot_longer(cols = 8:ncol(germ_update_planted_count_no_cum_germ),
               names_to= "Date",
               values_to = "Num_germ") %>%
  mutate(Date=mdy(Date))

#Convert date to days to germ
start_date <- as.Date("2025-04-24")

germ_long_dt_germ <- germ_long %>%
  mutate(Days_to_germ = as.integer(Date-start_date))
```

Germination rate per seed

```{r}
germ_long_dt_germ_rate <- germ_long_dt_germ %>%
  mutate(Germ_rate_seed = if_else(Num_germ == 1, 1/Days_to_germ, NA_real_),
         Germ_rate_seed = if_else(is.infinite(Germ_rate_seed), 1, Germ_rate_seed)) # 1 seeds germinated on the start day so change germ rate from inf to 1
```

Germination proportion by pot by day (is this different from germination prob? - no)

```{r}
germ_prop <- germ_update_planted_count %>%
  mutate(Germ_prop = Num_germ/seeds_per_pot)
```

Calculate average and SE germ prop and rate from pot proportions - probably not how I should do this

```{r}
avg_germ_prop <- germ_prop %>%
  group_by(pop.id) %>%
  summarize(Avg_germ_prop = mean(Germ_prop, na.rm=TRUE),
            SD_germ_prop = sd(Germ_prop, na.rm=TRUE),
            n=sum(Germ_prop != 0), #number of pots with germ, not total number of germ
            SE_germ_prop= SD_germ_prop/sqrt(n),
    .groups = "drop") %>%
  mutate(
    Cross_Type = case_when(
      str_count(pop.id, "x") == 0 ~ "Parent",
      str_count(pop.id, "x") == 1 ~ "F1",
      str_count(pop.id, "x") == 2 ~ "BC1",
      str_count(pop.id, "x") == 3 ~ "F2",
      TRUE ~ "Other"  # Fallback for unexpected cases
    )
  ) %>%
  ungroup()

total_planted <- germ_prop %>%
  group_by(pop.id) %>%
  summarise(total_seeds = sum(seeds_per_pot, na.rm=TRUE)) %>%
  ungroup()

avg_germ_prop <- avg_germ_prop %>%
  left_join(total_planted) %>%
  arrange(Cross_Type)%>%
  select(Cross_Type, pop.id, n, total_seeds, Avg_germ_prop, SD_germ_prop, SE_germ_prop)
```

Calculate average and SE germ rate from pot proportions - probably not how I should do this

```{r}
avg_germ_rate <- germ_long_dt_germ_rate %>%
  group_by(pop.id) %>%
  summarise(
    Avg_germ_rate_all = mean(Germ_rate_seed, na.rm=TRUE),
    SD_germ_rate_all = sd(Germ_rate_seed, na.rm=TRUE),
    n = sum(!is.na(Germ_rate_seed)),
    SE_germ_rate_all = SD_germ_rate_all/sqrt(n()),
    .groups="drop"
  ) %>%
  mutate(
    Cross_Type = case_when(
      str_count(pop.id, "x") == 0 ~ "Parent",
      str_count(pop.id, "x") == 1 ~ "F1",
      str_count(pop.id, "x") == 2 ~ "BC1",
      str_count(pop.id, "x") == 3 ~ "F2",
      TRUE ~ "Other"  # Fallback for unexpected cases
    )
  )
```

#### A better approach to estimate mean and var for germ prop - the averages are similar to taking the average of each pot proportion and the SE is smaller

```{r}
#pop.id as fixed effect
avg_prop <- glm(cbind(Num_germ, seeds_per_pot-Num_germ) ~ pop.id,
                data=germ_update_planted_count,
                family=binomial(link="logit"))

#intercept as a fixed effect to see if pop.id is significant
avg_prop_null <- glm(cbind(Num_germ, seeds_per_pot-Num_germ) ~ 1,
                data=germ_update_planted_count,
                family=binomial(link="logit"))

lrtest(avg_prop_null, avg_prop) #yes, pop.id is a significant predictor

summary(avg_prop)

estimates <- emmeans(avg_prop, ~pop.id, type="response") #using emmeans to extract avg and SE from the model, while also backtransforming it
summary(estimates)

#make a tibble and add cross type
estimates_tbl <- estimates_tbl %>%
  mutate(prob = format(prob, scientific = FALSE)) %>%
  mutate(SE = format(SE, scientific = FALSE)) %>%
  mutate(prob = as.numeric(prob),
         SE = as.numeric(SE)) %>%
  select(pop.id, prob, SE) %>%
  left_join(avg_germ_prop[,1:2])

#write csv
write.csv(estimates_tbl, file="WL2_2025_germ_prob.csv", row.names = FALSE)

#comparing the average estimates from the model to the average of the pot proportions - they look similar
test_comp <- estimates_tbl %>%
  left_join(avg_germ_prop, by="pop.id") %>%
  select(pop.id, Avg_germ_prop, prob,SE_germ_prop, SE)
```

```{r}
ggplot(estimates_tbl, aes(x=pop.id, y=prob)) +
  geom_point(size=0.9)+
  geom_errorbar(aes(ymin=prob-SE, ymax=prob+SE, width=0.1)) +
  facet_wrap(~Cross_Type, scales="free_x")+
  theme(axis.text.x=element_text(angle=90, hjust=1, size=5)) +
  ylim(0,1) +
  labs(y="Germination Proportion")
```

```{r}
#ggplot(avg_germ_rate, aes(x=pop.id, y=Avg_germ_rate_all)) + 
 # geom_point() +
  #geom_errorbar(aes(ymin=Avg_germ_rate_all-SE_germ_rate_all, ymax=Avg_germ_rate_all+SE_germ_rate_all, width=0.2)) +
 # theme(axis.text.x=element_text(angle=90, hjust=1, size=5)) +
  #ylim(0,1) +
  #facet_wrap(~Cross_Type, scales= "free_x")
```

```{r}
#ggplot(avg_germ_prop, aes(x=pop.id, y=Avg_germ_prop)) +
 # geom_point(size=0.9)+
  #geom_errorbar(aes(ymin=Avg_germ_prop-SE_germ_prop, ymax=Avg_germ_prop+SE_germ_prop, width=0.1)) +
 # facet_wrap(~Cross_Type, scales="free_x")+
#  theme(axis.text.x=element_text(angle=90, hjust=1, size=5)) +
 # ylim(0,1) +
  #labs(y="Total Germination Proportion")

```

```{r}
#germ_prop_summary_table <- avg_germ_prop %>%
 # arrange(Cross_Type, pop.id) %>%
  #kable(caption="Germination Prop Summary")
#print(germ_prop_summary_table)
```

```{r}
#print(as.data.frame(avg_germ_prop %>% filter(Cross_Type == "Parent")), row.names = FALSE)

#print(as.data.frame(avg_germ_prop %>% filter(Cross_Type == "F1")), row.names = FALSE)

#print(as.data.frame(avg_germ_prop %>% filter(Cross_Type == "F2")), row.names = FALSE)

#print(as.data.frame(avg_germ_prop %>% filter(Cross_Type == "BC1")), row.names = FALSE)
```

Code from Julin to calculate prop High elev parent

```{r}
el <- read_sheet("https://docs.google.com/spreadsheets/d/1FORIRL-1J15fD5iPqVyv_EWbc6uhV83-NL8SsjJL0Hw/edit#gid=0") %>% 
  filter(`Species Code`=="STTO") %>%
  mutate(elevation=unlist(`Elevation (m)`)) %>%
  select(pop.id=`Site code`, elevation) %>%
  mutate(pop.id=str_replace(pop.id, "YOSE", "YO"))

avg_germ_prop <- avg_germ_prop %>%
  mutate(pop.id = as.character(pop.id))
el <- el %>%
  mutate(pop.id = as.character(pop.id))

el <- el %>%
  mutate(high=elevation>2000)

avg_germ_prop_w_elev <- avg_germ_prop %>%
  left_join(el, by = "pop.id")
```

```{r}
high_regex <- el %>%
  filter(high, !is.na(pop.id)) %>% 
  pull(pop.id) %>%
  str_c(collapse = "|")

low_regex <- el %>%
  filter(!high, !is.na(pop.id)) %>% 
  pull(pop.id) %>%
  str_c(collapse = "|")

all_regex <- el %>%
  filter(!is.na(pop.id)) %>% 
  pull(pop.id) %>%
  str_c(collapse = "|")

```

```{r}
F1_avg_germ_prop <- avg_germ_prop_w_elev %>%
  filter(Cross_Type == "F1") %>%
  mutate(count.high=str_count(pop.id, high_regex),
         count.low=str_count(pop.id, low_regex),
         prop.high = count.high / (count.high + count.low),
         maternal.pop = str_extract(pop.id, "^[A-Z]{2,3}[1-9]{0,2}"),
         paternal.pop = str_extract(pop.id, "[A-Z]{2,3}[1-9]{0,2}$"),
         maternal.high=str_count(maternal.pop, high_regex),
         paternal.high=str_count(paternal.pop, high_regex)
  )
```

```{r}
F2_avg_germ_prop <- avg_germ_prop_w_elev %>%
  filter(Cross_Type == "F2") %>%
  mutate(
    maternal.pop = str_extract(pop.id, "^\\(.*?\\)|[A-Z]{2,3}[1-9]{0,2}"),
    paternal.pop = str_extract(pop.id, " \\(.*?\\)$|[A-Z]{2,3}[1-9]{0,2}$") %>% trimws(),
    maternal.grandmother.pop = str_extract(maternal.pop, "[A-Z]{2,3}[1-9]{0,2}"),
    maternal.grandmother.high = str_count(maternal.grandmother.pop, high_regex)
  ) %>%
  rowwise() %>%
  mutate(
    maternal.high = str_count(maternal.pop, high_regex) / str_count(maternal.pop, all_regex),
    paternal.high = str_count(paternal.pop, high_regex) / str_count(paternal.pop, all_regex),
    prop.high = mean(c(maternal.high, paternal.high), na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
cross_type <- avg_germ_prop[,c(1,6)]

germ_update_planted_count_cross_type <- germ_update_planted_count %>%
  left_join(cross_type, by="pop.id")

summary_table <- germ_update_planted_count_cross_type %>%
  group_by(Cross_Type,pop.id) %>%
  summarise(
    Pot_num = n(),
    Num_seeds_planted = sum(seeds_per_pot, na.rm = TRUE),
    Num_seeds_germ = sum(Num_germ, na.rm = TRUE),
    Num_new_pots = pmax(Num_seeds_germ - Pot_num, 0),
    .groups = "drop"
  )

write.csv(summary_table, "germ_summary_table.csv", row.names=FALSE)
print(summary_table, n=Inf)
```
